---
title: "How does covr work anyway?"
author: "Jim Hester"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How does covr work anyway?}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Introduction ##
This vignette walks you through the design of covr and breaks down the process
of tracking code execution piece by piece.

# Other coverage tools ##
Prior to writing covr, there were a handful of coverage tools for R code.
[R-coverage](http://r2d2.quartzbio.com/posts/r-coverage-docker.html) by Karl Forner and
[testCoverage](https://github.com/MangoTheCat/testCoverage) by Tom Taverner, Chris Campbell, Suchen Jin were the two
I was most aware of.

## R-coverage ###
`R-coverage` provided a very robust solution by modifying
the R source code to instrument the code for each call.  Unfortunately this
requires you to patch the R source and getting the changes upstreamed into the
base R distribution would likely be challenging.

## Test Coverage ###
`testCoverage` uses an alternate parser of R-3.0 to instrument R
code and record whether the code is run by tests. The package replaces symbols
in the code to be tested with a unique identifier. This is then injected into a
tracing function that will report each time the symbol is called. The first
symbol at each level of the expression tree is traced, allowing the coverage of
code branches to be checked.  This is a complicated implementation I do not fully
understand, which is one of the reasons I chose to write `covr`.

## Covr ###
Covr takes an approach in-between the two previous tools, modifying the
function definitions by parsing the abstract syntax tree and inserting trace
statements. These modified definitions are then transparently replaced in-place
using C. This allows us to correctly instrument every call and function in a
package without having to resort to alternate parsing or changes to the R
source.

# Modifying the call tree ##
The core function in covr is
[R/trace_calls.R](https://github.com/jimhester/covr/blob/59e17a5317a0fd69a6701e19317f20e27b6c44f5/R/trace_calls.R#L9-L72).
This function was adapted from
[pryr::modify_lang](https://github.com/hadley/pryr/blob/2ba41e433a59025267570247b933112009691c7a/R/modify-lang.r#L21-L49).
This recursive function modifies each of the leaves (atomic or name objects) of
a R expression by applying a given function to them.  For non-leaves we simply
call modify_lang recursively in various ways.  The logic behind modify_lang and
similar functions to parse and modify R's AST is explained in more detail at
[Walking the AST with recursive
functions](http://adv-r.had.co.nz/Expressions.html).

```{r eval = FALSE}
modify_lang <- function(x, f, ...) {
  recurse <- function(y) {
    # if (!is.null(names(y))) names(y) <- f2(names(y))
    lapply(y, modify_lang, f = f, ...)
  }

  if (is.atomic(x) || is.name(x)) {
    # Leaf
    f(x, ...)
  } else if (is.call(x)) {
    as.call(recurse(x))
  } else if (is.function(x)) {
    formals(x) <- modify_lang(formals(x), f, ...)
    body(x) <- modify_lang(body(x), f, ...)
    x
  } else if (is.pairlist(x)) {
    # Formal argument lists (when creating functions)
    as.pairlist(recurse(x))
  } else if (is.expression(x)) {
    # shouldn't occur inside tree, but might be useful top-level
    as.expression(recurse(x))
  } else if (is.list(x)) {
    # shouldn't occur inside tree, but might be useful top-level
    recurse(x)
  } else {
    stop("Unknown language class: ", paste(class(x), collapse = "/"),
      call. = FALSE)
  }
}
```

We can use this same framework to instead insert a trace statement before each
call by replacing each call with a call to a counting function followed by the
previous call.  Braces (`{`) in R may seem like language syntax, but
they are actually a Primitive function and you can call them like any other
function.

```{r}
identical({1+2;3+4}, `{`(1+2, 3+4))
```
Remembering that braces always return the value of the last evaluated
expression we can call a counting function followed by the previous function
substituting `as.call(recurse(x))` in our function above with.
```{r, eval = FALSE}
`{`(count(), as.call(recurse(x)))
```

## Source References ###
Now that we have a way to add a counting function to any call in the AST we
need a way to determine where in the code source that function came from.
Luckily for us R has a built-in method to provide this information in the form
of source references.  When `option(keep.source = TRUE)` (the default for
interactive sessions), a reference to the source code for functions is stored
along with the function definition.  This reference is used to provide the
original formatting and comments for the given function source.  In particular
each call in a function contains a `srcref` attribute, which can then be used
as a key to count just that call.

The actual source for `trace_calls` is slightly more complicated because we
want to initialize the counter for each call while we are walking the AST and
there are a few non-calls we also want to count.

# Replacing ##
After we have our modified function definition how do we re-define the function
to use the updated definition, and ensure that all other functions which call
the old function also use the new definition?  You might try redefining the function directly.

```{r}
f1 <- function() 1

f1 <- function() 2
f1() == 2
```

While this does work for the simple case of calling the new function in the
same environment, it fails if the another function calls a function in a
different environment.
```{r}
env <- new.env()
f1 <- function() 1
env$f2 <- function() f1() + 1

env$f1 <- function() 2

env$f2() == 3
```
As modifying external environments and correctly restoring them can be tricky
to get correct, we use the C function
[reassign_function](https://github.com/jimhester/covr/blob/9753e0e257b053059b85be90ef6eb614a5af9bba/src/reassign.c#L7-L20),
which is used in `testthat::with_mock()`.  This function takes a function name,
environment, old definition, new definition and copies the formals, body,
attributes and environment from the old function to the new function.  This
allows you to do an in-place replacement of a given function with a new
function and ensure that all references to the old function will use the new
definition.

# S4 classes ##
R's S3 and RC object oriented classes simply define functions directly in the packages
namespace, so they can be treated the same as any other function.  S4 methods
have a more complicated implementation where the function definitions are
placed in an enclosing environment based on the generic method they implement.
This makes getting the function definition more complicated.
```{r eval = FALSE}
replacements_S4 <- function(env) {
  generics <- getGenerics(env)

  unlist(recursive = FALSE,
    Map(generics@.Data, generics@package, USE.NAMES = FALSE,
      f = function(name, package) {
      what <- methodsPackageMetaName("T", paste(name, package, sep = ":"))

      table <- get(what, envir = env)

      lapply(ls(table, all.names = TRUE), replacement, env = table)
    })
  )
}
```
`replacements_S4` first gets all the generic functions for the package
environment.  Then for each generic function if finds the mangled meta package
name and gets the corresponding environment from the base environment.  All of
the functions within this environment are then traced.

# Compiled code ##
## Gcov ###
Test coverage of compiled code uses a completely different mechanism than that
of R code.  Fortunately we can take advantage of
[Gcov](https://gcc.gnu.org/onlinedocs/gcc-4.1.2/gcc/Gcov.html#Gcov), the
built-in coverage tool for [gcc](https://gcc.gnu.org/) and compatible reports
from [clang](http://clang.llvm.org/) versions 3.5 and greater.

Both of these compilers track execution coverage when given the
`-fprofile-arcs -ftest-coverage` flags.  In addition it is necessary to turn
off compiler optimization `-O0`, otherwise the coverage output is difficult or
impossible to interpret as multiple lines can be optimized into one, functions
can be inlined ect.

## Makevars ###
R passes flags defined in `PKG_CFLAGS` to the compiler, however it also has
default flags including `-02` (defined in `$R_HOME/etc/Makeconf`) which need to
be overridden. Unfortunately it is not possible to override the default flags
with environment variables (as the new flags are added to the left of the
defaults rather than the right).  However if Make variables are defined in
`~/.R/Makevars` they _are_ used in place of the defaults.

Therefore we need to temporarily add `-O0 -fprofile-arcs -ftest-coverage` to
the Makevars file, then restore the previous state after the coverage is run.
The implementation of this is in
[R/makevars.R](https://github.com/jimhester/covr/blob/9753e0e257b053059b85be90ef6eb614a5af9bba/R/makevars.R).

## Subprocess ###
The last hurdle to getting compiled code coverage working properly is that the
coverage output is only produced when the running process ends.  Therefore you
cannot run the tests and get the results in the same R process.  In order to
handle this situation we use a [subprocess function](R/subprocess).  This
function aims to replicate as much of the calling environment as possible, then
calls the given expressions in that environment and returns any objects created
back to the calling environment.  In this way we can have a fully transparent
subprocess that acts like running the same code in the current
process.

## Compilation ###
The package's code is compiled by calling `devtools::load_all()` with
`recompile=TRUE` so while it does take some time to recompile all the code with
tracing enabled, you do not have to do a full `R CMD build` cycle.

# Running Tests ##
R code defined in `tests` or `inst/tests` is run using
`testthat::source_dir()`, which simply sources each of the R files in those
directories.  This makes `covr` compatible with any testing framework.

# Output Formats ##
The output format returned by covr is a very simple one.  It consists of a
named character vector, where the names are colon delimited information from
the source references.  Namely the file, line and columns the traced call is
from. The value is simply the number of times that given call was called. There
is also an `as.data.frame` method to make subsetting by various features simple.

While `covr` traces coverage by expression, typically users expect coverage to
be reported by line.

# Coveralls.io ##
Coveralls is a web service to help you track your code coverage over time, and
ensure that all your new code is fully covered.

The [coveralls API](https://coveralls.zendesk.com/hc/en-us/articles/201774865-API-Introduction)
is fairly simple assuming you are already building on a CI service like travis.
You send a JSON file to coveralls with the `service_job_id` (environment
variable TRAVIS_JOB_ID), the service name (`travis-ci`) and a list of source
files; each with the file name, file content and coverage per line for that file.

Once this JSON file is sent to the coveralls.io service it automatically
generates coverage reports for the current state and tracks changes in test
coverage over time.

# Conclusion ##
Covr aims to be a simple easy to understand implementation, hopefully this
vignette helps to explain the rational of the package and explain how and why it works.
