---
title: "How does covr work anyway?"
author: "Jim Hester"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{How does covr work anyway}
 %\VignetteEngine{knitr::rmarkdown}
 \usepackage[utf8]{inputenc}
---

# Introduction #
This vignette walks you through the design of covr and breaks down the process
of tracking code execution piece by piece.

# Other coverage tools #
Prior to writing covr, there were a handful of coverage tools for R code.
[R-coverage](http://r2d2.quartzbio.com/posts/r-coverage-docker.html) by Karl Forner and
[testCoverage](https://github.com/MangoTheCat/testCoverage) by Tom Taverner, Chris Campbell, Suchen Jin were the two
I was most aware of.

## R-coverage ##
`R-coverage` provided a very robust solution by modifying
the R source code to instrument the code for each call. Unfortunately this
requires you to patch the R source and getting the changes upstreamed into the
base R distribution would likely be challenging.

## Test Coverage ##
`testCoverage` uses an alternate parser of R-3.0 to instrument R
code and record whether the code is run by tests. The package replaces symbols
in the code to be tested with a unique identifier. This is then injected into a
tracing function that will report each time the symbol is called. The first
symbol at each level of the expression tree is traced, allowing the coverage of
code branches to be checked. This is a complicated implementation I do not fully
understand, which is one of the reasons I decided to write `covr`.

## Covr ##
Covr takes an approach in-between the two previous tools, modifying the
function definitions by parsing the abstract syntax tree and inserting trace
statements. These modified definitions are then transparently replaced in-place
using C. This allows us to correctly instrument every call and function in a
package without having to resort to alternate parsing or changes to the R
source.

# Modifying the call tree ##
The core function in covr is
[R/trace_calls.R](https://github.com/jimhester/covr/blob/59e17a5317a0fd69a6701e19317f20e27b6c44f5/R/trace_calls.R#L9-L72).
This function was adapted from
[pryr::modify_lang](https://github.com/hadley/pryr/blob/2ba41e433a59025267570247b933112009691c7a/R/modify-lang.r#L21-L49).
This recursive function modifies each of the leaves (atomic or name objects) of
a R expression by applying a given function to them. For non-leaves we simply
call `modify_lang` recursively in various ways. The logic behind `modify_lang` and
similar functions to parse and modify R's AST is explained in more detail at
[Walking the AST with recursive
functions](http://adv-r.had.co.nz/Expressions.html).

```{r eval = FALSE}
modify_lang <- function(x, f, ...) {
 recurse <- function(y) {
  # if (!is.null(names(y))) names(y) <- f2(names(y))
  lapply(y, modify_lang, f = f, ...)
 }

 if (is.atomic(x) || is.name(x)) {
  # Leaf
  f(x, ...)
 } else if (is.call(x)) {
  as.call(recurse(x))
 } else if (is.function(x)) {
  formals(x) <- modify_lang(formals(x), f, ...)
  body(x) <- modify_lang(body(x), f, ...)
  x
 } else if (is.pairlist(x)) {
  # Formal argument lists (when creating functions)
  as.pairlist(recurse(x))
 } else if (is.expression(x)) {
  # shouldn't occur inside tree, but might be useful top-level
  as.expression(recurse(x))
 } else if (is.list(x)) {
  # shouldn't occur inside tree, but might be useful top-level
  recurse(x)
 } else {
  stop("Unknown language class: ", paste(class(x), collapse = "/"),
   call. = FALSE)
 }
}
```

We can use this same framework to instead insert a trace statement before each
call by replacing each call with a call to a counting function followed by the
previous call. Braces (`{`) in R may seem like language syntax, but
they are actually a Primitive function and you can call them like any other
function.

```{r}
identical({ 1 + 2; 3 + 4 }, `{`(1 + 2, 3 + 4))
```
Remembering that braces always return the value of the last evaluated
expression we can call a counting function followed by the previous function
substituting `as.call(recurse(x))` in our function above with.
```{r, eval = FALSE}
`{`(count(), as.call(recurse(x)))
```

## Source References ##
Now that we have a way to add a counting function to any call in the AST
without changing the output we need a way to determine where in the code source
that function came from. Luckily for us R has a built-in method to provide this
information in the form of source references. When `option(keep.source =
TRUE)` (the default for interactive sessions), a reference to the source code
for functions is stored along with the function definition. This reference is
used to provide the original formatting and comments for the given function
source. In particular each call in a function contains a `srcref` attribute,
which can then be used as a key to count just that call.

The actual source for `trace_calls` is slightly more complicated because we
want to initialize the counter for each call while we are walking the AST and
there are a few non-calls we also want to count.


## Refining Source References ##
Each statement comes with a source reference. Unfortunately, the following is
counted as one statement:

```r
if (x)
 y()
```

To work around this, detailed parse data (obtained from a refined version of
`getParseData()`) is analyzed to impute source references at sub-statement
level for `if`, `for` and `while` constructs.

# Replacing #
After we have our modified function definition how do we re-define the function
to use the updated definition, and ensure that all other functions which call
the old function also use the new definition? You might try redefining the function directly.

```{r}
f1 <- function() 1

f1 <- function() 2
f1() == 2
```

While this does work for the simple case of calling the new function in the
same environment, it fails if the another function calls a function in a
different environment.
```{r}
env <- new.env()
f1 <- function() 1
env$f2 <- function() f1() + 1

env$f1 <- function() 2

env$f2() == 3
```
As modifying external environments and correctly restoring them can be tricky
to get correct, we use the C function
[reassign_function](https://github.com/jimhester/covr/blob/9753e0e257b053059b85be90ef6eb614a5af9bba/src/reassign.c#L7-L20),
which is used in `testthat::with_mock()`. This function takes a function name,
environment, old definition, new definition and copies the formals, body,
attributes and environment from the old function to the new function. This
allows you to do an in-place replacement of a given function with a new
function and ensure that all references to the old function will use the new
definition.

# S4 classes #
R's S3 object oriented classes simply define functions directly in the packages
namespace, so they can be treated the same as any other function. S4 methods
have a more complicated implementation where the function definitions are
placed in an enclosing environment based on the generic method they implement.
This makes getting the function definition more complicated.
```{r eval = FALSE}
replacements_S4 <- function(env) {
 generics <- getGenerics(env)

 unlist(recursive = FALSE,
  Map(generics@.Data, generics@package, USE.NAMES = FALSE,
   f = function(name, package) {
   what <- methodsPackageMetaName("T", paste(name, package, sep = ":"))

   table <- get(what, envir = env)

   lapply(ls(table, all.names = TRUE), replacement, env = table)
  })
 )
}
```
`replacements_S4` first gets all the generic functions for the package
environment. Then for each generic function if finds the mangled meta package
name and gets the corresponding environment from the base environment. All of
the functions within this environment are then traced.

# RC classes #
Similarly to S4 classes reference class (RC) classes define their methods in a
special environment. A similar method is used to add the tracing calls to the
class definition. These calls are then copied to the object methods when the
generator function is run.

# Compiled code #
## Gcov ##
Test coverage of compiled code uses a completely different mechanism than that
of R code. Fortunately we can take advantage of
[Gcov](https://gcc.gnu.org/onlinedocs/gcc-4.1.2/gcc/Gcov.html#Gcov), the
built-in coverage tool for [gcc](https://gcc.gnu.org/) and compatible reports
from [clang](http://clang.llvm.org/) versions 3.5 and greater.

Both of these compilers track execution coverage when given the `--coverage`
flag. In addition it is necessary to turn off compiler optimization `-O0`,
otherwise the coverage output is difficult or impossible to interpret as
multiple lines can be optimized into one, functions can be inlined etc.

## Makevars ##
R passes flags defined in `PKG_CFLAGS` to the compiler, however it also has
default flags including `-02` (defined in `$R_HOME/etc/Makeconf`) which need to
be overridden. Unfortunately it is not possible to override the default flags
with environment variables (as the new flags are added to the left of the
defaults rather than the right). However if Make variables are defined in
`~/.R/Makevars` they _are_ used in place of the defaults.

Therefore we need to temporarily add `-O0 --coverage` to
the Makevars file, then restore the previous state after the coverage is run.
The implementation of this is in
[R/makevars.R](https://github.com/jimhester/covr/blob/9753e0e257b053059b85be90ef6eb614a5af9bba/R/makevars.R).

## Subprocess ##
The last hurdle to getting compiled code coverage working properly is that the
coverage output is only produced when the running process ends. Therefore you
cannot run the tests and get the results in the same R process.
`tests::testPackages()` runs a separate R process when running tests, however
we need to modify the package code first before running the tests.

The current behavior in covr is to install the package to be tested in a
temporary directory, then add calls to the lazy loading code to install a user
hook which modifies the code when it is loaded. We also register a finalizer
which prints the coverage couts when the namespace is unloaded or the R process
exists. These output files are then aggregated together to determine the
coverage.

This procedure works regardless of the number of child R processes used, so
therefore also works with parallel code.

# Output Formats #
The output format returned by covr is a very simple one. It consists of a
named character vector, where the names are colon delimited information from
the source references. Namely the file, line and columns the traced call is
from. The value is simply the number of times that given call was called and
the source ref of the original call. There is also an `as.data.frame` method to
make subsetting by various features easy to do

While `covr` tracks coverage by expression, typically users expect coverage to
be reported by line, so there are functions to convert to line oriented
coverage.

# Codecov.io and Coveralls.io #
Codecov and Coveralls are a web services to help you track your code coverage
over time, and ensure that all your new code is fully covered.

They both have simple JSON based APIs to submit and report on coverage.

# Conclusion #
Covr aims to be a simple easy to understand implementation, hopefully this
vignette helps to explain the rationale behind the package and explain how and
why it works.
